Distortion-corrected kernel density estimate on Riemannian manifolds

### 2022-10-10 paper/paper.Rmd
100D simulation rank plot

For $\pmb{p} \in M$, denote an open ball centered at $\pmb{p}$ with radius $r$ as $B(\pmb{p},r)$ and the open neighborhood of $0_{\pmb{p}} \in T_{\pmb{p}}M$ can be denoted as $B(0_{\pmb{p}},r)=\exp_{\pmb{p}}^{-1}(B(\pmb{p},r))$. Consider the Riemannian metrics $g^{\prime}$ and $g^{\prime \prime}$ in $B(0_{\pmb{p}},r)$, with $g^{\prime} = \exp^{\ast}_{\pmb{p}}(g)$ being the pullback of $g$ induced by the exponential map and $g^{\prime \prime}$ being the canonical metric induced by $g_{\pmb{p}}$ in $B(0_{\pmb{p}},r)$. 
Let $u \in B(0_{\pmb{p}},r)$ be a tangent vector and for any given chart $(\bar{U}, \bar{\psi})$ of $B(0_{\pmb{p}},r)$, we have $u \in \bar{U}$. 
The volumes of the parallelepiped spanned by the basis $\{(\partial / \partial\bar{\psi}_1|_u), \dots, (\partial / \partial\bar{\psi}_d|_u)\}$ are given by $| \det g^{\prime}((\partial / \partial\bar{\psi}_i|_u), (\partial / \partial\bar{\psi}_j|_u)) |^{1/2}$ with respect to $g^{\prime}$, and $| \det g^{\prime \prime}((\partial / \partial\bar{\psi}_i|_u), (\partial / \partial\bar{\psi}_j|_u)) |^{1/2}$ with respect to $g^{\prime \prime}$. The quotient of these two volumes is then independent of the chart $(\bar{U}, \bar{\psi})$ [@Henry2009-om]. 
Therefore, given a point $\pmb{q} \in B(\pmb{p},r)$ and a tangent vector $u = \exp^{-1}_{\pmb{p}}(\pmb{q}) \in B(0_{\pmb{p}},r)$, for any chart $(\bar{U}, \bar{\psi})$ of $B(0_{\pmb{p}},r)$ that contains $u$, @Henry2009-om defines the volume density function on the Riemannian manifold $(M,g)$ as
\begin{equation}
\label{eq:volumedensityratio}
\theta_p(q) = \frac{| \det g^{\prime}((\partial / \partial\bar{\psi}_i|_u), (\partial / \partial\bar{\psi}_j|_u)) |^{1/2} }{ | \det g^{\prime \prime}((\partial / \partial\bar{\psi}_i|_u), (\partial / \partial\bar{\psi}_j|_u)) |^{1/2} } \\ .
\end{equation}
The volume density function is only defined for points satisfying $d_g(\pmb{p},\pmb{q}) < inj_gM$ so that there exists an open ball $B(\pmb{p},r) \in T_{\pmb{p}}M$. And $\theta_{\pmb{p}}(\pmb{q})=1$ for all $\pmb{p},\pmb{q} \in \mathbb{R}^d$ when the manifold $M$ is $\mathbb{R}^d$ with the canonical metric.

@@

```{r fivedisomapden, fig.align = 'center', out.width = "95%", fig.cap="Scatterplot of true density and estimated density of different embeddings for KDE with both variable and fixed bandwidth. The point shapes indicates the two Gaussian mixture components and the colors shows the distance to the kernel cores.", echo = FALSE}
# knitr::include_graphics("figures/fived_density_comparison_4ml_riem20.png")
knitr::include_graphics("figures/sim4d10000_density_comparison_4ml_riem1k200_rawdensity.png")
```

@@

```{r fivedcors, echo=FALSE, message=FALSE, eval=T}
load("figures/CorrelationTable_fived_4ml_riem20.rda")
col.max <- apply(cors, 2, max)
max.match <- rep(col.max, each = 2) == cors
data.frame(cors) %>% 
  mutate_if(is.numeric, format, digits = 3, nsmall = 3) %>% 
  kableExtra::kbl(caption = "Correlation between true density and estimated density for four manifold learning embeddings.", booktabs = TRUE, digits = 3, escape = FALSE) %>%
  # kable_styling(latex_options = "scale_down") %>% # automatically adjust the table to page width
  kable_paper(full_width = FALSE) %>%
  # column_spec(1, width = "6cm") %>% 
  column_spec(2, bold = max.match[,1]) %>%
  column_spec(3, bold = max.match[,2]) %>%
  column_spec(4, bold = max.match[,3]) %>%
  column_spec(5, bold = max.match[,4])
```

@@

The CDF of X is given by
$$
P(X \leq x) = P(\cos(\theta) \leq x) \\
= P(\theta \geq \arccos(x)) \\
= \int_{\arccos(x)}^{\pi/2} f(\theta) d\theta \\
=  \frac{2}{\pi} (\frac{\pi}{2} - \arccos(x) ) \\
= 1 - \frac{2}{\pi} \arccos(x).
$$
Then the density of X is the derivative of the CDF w.r.t. $x$ $$
f(x) = -\frac{2}{\pi} (-\frac{1}{\sqrt{1 - x^2}}) \\
= \frac{2}{\pi \sqrt{1 - x^2}}.
$$

@@

Let $S^2$ be the two-dimensional sphere of radius 1 and $p \in S^2$. Let $v$ and $w$ be vectors such that $\|v\| = \|w\| = 1$ and $\{v, w, p\}$ is the orthonormal basis of $\mathbb{R}^3$. Consider the exponential chart $(U, \Psi)$ induced by the parametrisation $\Psi^{-1}: B_\pi(0) \rightarrow S^2 \backslash \{-p\}$ given by
$$
\Psi^{-1}(s,t) = \cos(r)p + \frac{\sin(r)}{r}(sv+tw),
$$
if $(s,t) \neq (0,0)$ and $\Psi^{-1}(0,0)=p$ where $r=\sqrt{s^2 + t^2}$.
Note that $r=d_g(p,q)$ if $q = \exp_p(sv+tw)$.
Then we take the partial derivatives of $\Psi^{-1}(s,t)$ with respect to the basis of the tangent space, $s$ and $t$, and have that
$$
A(s,t) = \frac{\partial}{\partial \Psi_1} \bigg|_{\Psi^{-1}(s,t)} = \frac{-\sin{(r)}s}{r}p + \frac{\cos(r)rs^2+t^2\sin(r)}{r^3}v + \frac{(\cos(r)r-\sin(r))st}{r^3}w,
$$
$$
B(s,t) = \frac{\partial}{\partial \Psi_2} \bigg|_{\Psi^{-1}(s,t)} = \frac{-\sin{(r)}t}{r}p + \frac{(\cos(r)r-\sin(r))st}{r^3}v + \frac{\cos(r)rt^2+\sin(r)s^2}{r^3}w.
$$
The coefficients for $v$ and $w$ gives the local coordinates of two tangent vectors.
Further, since the riemannian metric is defined as the inner product between two tangent vectors, each component of the riemannian metric could be calculated as $g(A(s,t), A(s,t)), g(A(s,t), B(s,t)), g(B(s,t), B(s,t))$.
Therefore, we obtain that the volume density on the sphere is
$$
\theta_p(q) = \bigg| \det g_q\bigg( \frac{\partial}{\partial \Psi_i}\bigg|_q, \frac{\partial}{\partial \Psi_j}\bigg|_q \bigg) \bigg|^{1/2} \\
= \frac{|\sin(d_g(p,q))|}{d_g(p,g)},
$$
for $q \neq p, -p$, and $\theta_p(p) = 1$.

>>>>

For $\pmb{p} \in M$, denote an open ball centered at $\pmb{p}$ with radius $r$ as $B(\pmb{p},r)$ and the open neighborhood of $0_{\pmb{p}} \in T_{\pmb{p}}M$ can be denoted as $B(0_{\pmb{p}},r)=\exp_{\pmb{p}}^{-1}(B(\pmb{p},r))$. Consider the Riemannian metrics $g^{\prime}$ and $g^{\prime \prime}$ in $B(0_{\pmb{p}},r)$, with $g^{\prime} = \exp^{\ast}_{\pmb{p}}(g)$ being the pullback of $g$ induced by the exponential map and $g^{\prime \prime}$ being the canonical metric induced by $g_{\pmb{p}}$ in $B(0_{\pmb{p}},r)$. 
Let $u \in B(0_{\pmb{p}},r)$ be a tangent vector and for any given chart $(\bar{U}, \bar{\psi})$ of $B(0_{\pmb{p}},r)$, we have $u \in \bar{U}$. 
The volumes of the parallelepiped spanned by the basis $\{(\partial / \partial\bar{\psi}_1|_u), \dots, (\partial / \partial\bar{\psi}_d|_u)\}$ are given by $| \det g^{\prime}((\partial / \partial\bar{\psi}_i|_u), (\partial / \partial\bar{\psi}_j|_u)) |^{1/2}$ with respect to $g^{\prime}$, and $| \det g^{\prime \prime}((\partial / \partial\bar{\psi}_i|_u), (\partial / \partial\bar{\psi}_j|_u)) |^{1/2}$ with respect to $g^{\prime \prime}$. The quotient of these two volumes is then independent of the chart $(\bar{U}, \bar{\psi})$ [@Henry2009-om]. 
Therefore, given a point $\pmb{q} \in B(\pmb{p},r)$ and a tangent vector $u = \exp^{-1}_{\pmb{p}}(\pmb{q}) \in B(0_{\pmb{p}},r)$, for any chart $(\bar{U}, \bar{\psi})$ of $B(0_{\pmb{p}},r)$ that contains $u$, @Henry2009-om defines the volume density function on the Riemannian manifold $(M,g)$ as
\begin{equation}
\label{eq:volumedensityratio}
\theta_p(q) = \frac{| \det g^{\prime}((\partial / \partial\bar{\psi}_i|_u), (\partial / \partial\bar{\psi}_j|_u)) |^{1/2} }{ | \det g^{\prime \prime}((\partial / \partial\bar{\psi}_i|_u), (\partial / \partial\bar{\psi}_j|_u)) |^{1/2} }  .
\end{equation}
The volume density function is only defined for points satisfying $d_g(\pmb{p},\pmb{q}) < inj_gM$ so that there exists an open ball $B(\pmb{p},r) \in T_{\pmb{p}}M$. And $\theta_{\pmb{p}}(\pmb{q})=1$ for all $\pmb{p},\pmb{q} \in \mathbb{R}^d$ when the manifold $M$ is $\mathbb{R}^d$ with the canonical metric.

@@

```{r fivedisomapden, fig.align = 'center', out.width = "95%", fig.cap="Rank comparison between the true density and estimated density from both DC-KDE and KDE. Four manifold learning methods are used rowwise. The point shapes indicates whether they are the true outleirs, and the grey shading highlights the top 1\\% rank region. The colors show the distance to the center of the semisphere, with darker points being distant from the center.", echo = FALSE}
# knitr::include_graphics("figures/sim4d10000_density_comparison_4ml_riem1k200_rawdensity.png")
knitr::include_graphics("figures/sim4d10000_density_comparison_4ml_radius10_k200_rankdensity_circleoutlier_with1rec.png")
```

@@

```{r fivedcors, echo=FALSE, message=FALSE, eval=T}
load("figures/CorrelationTable_4d_N10000_4ml_radius10_r1.rda")
col.max <- apply(cors, 2, max)
max.match <- rep(col.max, each = 2) == cors
data.frame(cors) %>% 
  mutate_if(is.numeric, format, digits = 3, nsmall = 3) %>% 
  kableExtra::kbl(caption = "Correlation between true density and estimated density for four manifold learning embeddings.", booktabs = TRUE, digits = 3, escape = FALSE) %>%
  # kable_styling(latex_options = "scale_down") %>% # automatically adjust the table to page width
  kable_paper(full_width = FALSE) %>%
  # column_spec(1, width = "6cm") %>% 
  column_spec(2, bold = max.match[,1]) %>%
  column_spec(3, bold = max.match[,2]) %>%
  column_spec(4, bold = max.match[,3]) %>%
  column_spec(5, bold = max.match[,4])
```

@@

The CDF of X is given by
$$
\begin{aligned}
P(X \leq x) =& P(\cos(\theta) \leq x) \\
=& P(\theta \geq \arccos(x)) \\
=& \int_{\arccos(x)}^{\pi/2} f(\theta) d\theta \\
=&  \frac{2}{\pi} (\frac{\pi}{2} - \arccos(x) ) \\
=& 1 - \frac{2}{\pi} \arccos(x).
\end{aligned}
$$
Then the density of X is the derivative of the CDF w.r.t. $x$ 
$$
\begin{aligned}
f(x) =& -\frac{2}{\pi} (-\frac{1}{\sqrt{1 - x^2}}) \\
=& \frac{2}{\pi \sqrt{1 - x^2}}.
\end{aligned}
$$

@@

Let $S^2$ be the two-dimensional sphere of radius 1 and $p \in S^2$. Let $v$ and $w$ be vectors such that $\|v\| = \|w\| = 1$ and $\{v, w, p\}$ is the orthonormal basis of $\mathbb{R}^3$. Consider the exponential chart $(U, \Psi)$ induced by the parametrisation $\Psi^{-1}: B_\pi(0) \rightarrow S^2 \backslash \{-p\}$ given by
$$
\Psi^{-1}(s,t) = \cos(r)p + \frac{\sin(r)}{r}(sv+tw),
$$
if $(s,t) \neq (0,0)$ and $\Psi^{-1}(0,0)=p$ where $r=\sqrt{s^2 + t^2}$.
Note that $r=d_g(p,q)$ if $q = \exp_p(sv+tw)$.
Then we take the partial derivatives of $\Psi^{-1}(s,t)$ with respect to the basis of the tangent space, $s$ and $t$, and have that
$$
A(s,t) = \frac{\partial}{\partial \Psi_1} \bigg|_{\Psi^{-1}(s,t)} = \frac{-\sin{(r)}s}{r}p + \frac{\cos(r)rs^2+t^2\sin(r)}{r^3}v + \frac{(\cos(r)r-\sin(r))st}{r^3}w,
$$
$$
B(s,t) = \frac{\partial}{\partial \Psi_2} \bigg|_{\Psi^{-1}(s,t)} = \frac{-\sin{(r)}t}{r}p + \frac{(\cos(r)r-\sin(r))st}{r^3}v + \frac{\cos(r)rt^2+\sin(r)s^2}{r^3}w.
$$
The coefficients for $v$ and $w$ gives the local coordinates of two tangent vectors.
Further, since the riemannian metric is defined as the inner product between two tangent vectors, each component of the riemannian metric could be calculated as $g(A(s,t), A(s,t)), g(A(s,t), B(s,t)), g(B(s,t), B(s,t))$.
Therefore, we obtain that the volume density on the sphere is
$$
\theta_p(q) = \bigg| \det g_q\bigg( \frac{\partial}{\partial \Psi_i}\bigg|_q, \frac{\partial}{\partial \Psi_j}\bigg|_q \bigg) \bigg|^{1/2} 
= \frac{|\sin(d_g(p,q))|}{d_g(p,g)},
$$
for $q \neq p, -p$, and $\theta_p(p) = 1$.
